---
title: 'LAB: Linear Regression'
date: "`r Sys.time()`"
draft: false 

execute:
  echo: true
  eval: true
  collapse: true


format:
  html:
    output-file: lab-linear-regression.html
  pdf:
    output-file: lab-linear-regression.pdf


---


{{< include _preamble.qmd >}}


# GSS dataset and `gssr` package 

```{r}
stopifnot(
  require(gssr),
  require(gssrdoc),
  require(tidyverse),
  require(lobstr),
  require(rlang)
)
```

## General social survey


## Exploring the dictionary  `gss_dict`


::: {.callout-note title="Question"}

- Get the schema of `gss_dict` from `gssrdoc`
- What is the class/type of columns `years` ?
- Extract elements `1..3` from column `years`
- Have a look at [`tidyr`](https://tidyr.tidyverse.org/articles/nest.html) documentation about *nested* data. 
- Which variables were used in 1982?  
:::

::: {.content-visible when-profile='solution'} 
 
::: {.callout-tip title="Solution"}
 
 ```{r}
gss_dict |>
  glimpse(width=50)
 ```


```{r}
base::class(gss_dict$years)
rlang::is_bare_list(gss_dict$years)
is.data.frame(gss_dict$years[[2]])
is.data.frame(gss_dict$years[[1]])
```

`gss_dict`  is a *nested* dataframe: column `years` (at least) is of class `list`.
Indeed `gss_dict$years` is a list, and some of its elements are dataframes. 


```{r}
gssrdoc::gss_dict[2:10, ] |> 
  dplyr::select(variable, years) |>
  unnest(years)
  mutate(across(everything(), str)) |>  

```



:::


::: {.callout-tip}

### 

:::
:::



::: {.callout-note title="Question"}

-  Extract the data corresponding to year 1982. 
-  Save the dataframe/tibble as a parquet file in the 'DATA' subdirectory.
-  The GSS data comes in a *labelled format*, what does it mean?

:::


::: {.content-visible when-profile='solution'} 

::: {.callout-tip title="Solutioj"}

```{r}
fpath <- './DATA/gss_1982.parquet'

if (!fs::file_exists(fpath)) {
  df <- gssr::gss_get_yr(1982)
  df |>
    arrow::write_parquet(fpath)
} else {
  df <- arrow::read_parquet(fpath)
}
```

::: 


::: {.callout-note title="Question"}

- Inspect documentation of `gssr`
- Which demographic variables were used during the 1982 survey?
- What is the difference between cross-sectional data  and panel data?
- How would you qualify the 1982 survey?
- Load `gssrdoc` to have online documentation 
- Request help about variable `wrkstat`
- Which are the numerical variables?
- Have a look at [`gssrdoc` reference page](https://kjhealy.github.io/gssrdoc/reference/index.html)
- Which column names contain `inc`?
- Which columns are numerical?

:::


::: {.content-visible when-profile='solution'} 
 
::: {.callout-tip title="Solution"}


```{r}
var_choice <- c(
  "sex",
  "age",
  "occ",
  "childs",
  "coninc",
  "income",
  "rincome",
  "income82",
  "income82",
  "size",
  "educ",
  "region",
  "marital",
  "famgen",
  "region",
  "respnum"
)

cat_vars <- c(
  "sex",
  "occ",
  "income",
  "rincome",
  "income82",
  "size",
  "educ",
  "region",
  "marital",
  "famgen",
  "region",
  "respnum"
)
```
```{r}
df <- df |>
  select(all_of(var_choice)) |>
  mutate(across(all_of(cat_vars), as_factor))  
```

```{r}
df |> 
  gt::gt()
```


```{r}
arrow::read_parquet(fpath, 
                    col_select=c(
    age,
    sex,
    wrkstat,
    hrs2,
    partyid,
    vote76,
    pres76,
    if76who,
    vote80, pres80, 
    if80who,
    polviews,
    starts_with("occ"),
    starts_with("wrk")   
  )) |>
  head()
```

```{r}

```
::: 
:::


## 

```{r}
df |>
  ggplot() +
  aes(x=age, y=coninc) + 
  geom_jitter(alpha=.5, size=.2) +
  scale_y_log10() +
  geom_smooth(method="lm")
```

```{r}
lm(coninc ~ age, df) |> summary()
```